Msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-07-27 paper.tex

We have implemented selective sweeps in \msprime\ similarly to
the \texttt{discoal} implementation-- sweep trajectories are
generated from a jump process approximation to the conditional diffusion
of an allele bound for fixation~\citep{coop2004ancestral}.
Details of our implementation are available in the Appendix.
This implementation allows us to further generalize sweep models to
changing population sizes as well as selection from standing variation,
both of which are features that are planned for the near future.
Moreover because this implements the basic infrastructure for the structured
coalescent, adding support for, e.g.,
inversions is straightforward~\citep{peischl2013sequential}.

@@

\label{app-sweeps}
\section*{Selective sweep trajectories}
We have implemented the coalescent with selective sweeps as a structured 
coalescent process (as in~\citep{braverman1995hitchhiking}). During the
``sweep phase'' of the simulation the rate of coalescence and recombination
is tracked separately in lineages linked to the beneficial allele or not,
with those rates depending on the frequency of the beneficial allele in 
the population at a given time. The allele frequency trajectory of the
sweeping allele is modeled using a jump process that approximates
the conditional diffusion of the allele, conditioning on its eventual 
fixation~\citep{coop2004ancestral}. 
The jump process moves back in time following the beneficial
allele frequency, $p$, from some initial frequency (e.g., $p=1$) back to
the origination of the allele at $p=1/2N$, and we track time in small
increments $\delta t$. This works such that the allele frequency at
the next time step back in time, call it $p^\prime$ is

\begin{eqnarray*}
	p'=\left\{ \begin{array}{ll}
	    p + \mu(p)\delta t +\sqrt{p(1-p)\delta t} & \quad \mbox{with probability $1/2$}\\
	    p + \mu(p)\delta t -\sqrt{p(1-p)\delta t} & \quad \mbox{with probability $1/2$}\\
	\end{array} \right.
\end{eqnarray*}
where $\alpha=2Ns$ and $s$ is the fitness advantage in homozygotes. This model
assumes genic selection (i.e., that the dominance coefficient $h=0.5$. Further
$\mu(p)$ is
\begin{eqnarray*}
	\mu(p)=\frac{\alpha p (1 - p)}{tanh(\alpha(1-p))}
\end{eqnarray*}

While our current implementation only deals with genic selection,
we plan in the future to add sweeps with arbitrary dominance. Further planned
additions include trajectories of purely neutral alleles, as well as trajectories
that model soft selective sweeps.

>>>>

We have implemented selective sweeps in \msprime\ similarly to
the \texttt{discoal} implementation-- sweep trajectories are
generated from a jump process approximation to the conditional diffusion
of an allele bound for fixation~\citep{coop2004ancestral}.
This implementation allows us to further generalize sweep models to
changing population sizes as well as selection from standing variation,
both of which are features that are planned for the near future.
Moreover because this implements the basic infrastructure for the structured
coalescent, adding support for, e.g.,
inversions is straightforward~\citep{peischl2013sequential}.

@@
